### 2021-03-09 paper/rebuttal/rebuttal.tex

\subsection{\textit{word2vec} algorithm}
Imagine a trajectory, denoted by ($a_{1}, a_{2}, \ldots, a_{T}$), where $a_{t}$ is the $t$th location in the trajectory. A location, $a_{t}$, is considered to have context locations, $a_{t-w}, \ldots, a_{t-1}, a_{t+1},\ldots, a_{t+w}$, that appear in the window surrounding $a_t$ up to a time lag of $w$, where $w$ is the window size parameter truncated at $t - w \geq 0$ and $t + w \leq T$.
Then, the skip-gram \textit{word2vec} learns probability $p(a_{t + \tau} \vert a_{t})$, where $-w\leq \tau\leq w$ and $\tau \neq 0$,  by maximizing its log likelihood given by
%
% MAIN W2V Equation
%
\begin{equation}
	{\cal J} = \frac{1}{T}\sum_{t = 1}^{T} \sum_{-w \leq \tau \leq w, \tau \neq 0} \log P(a_{t + \tau} \vert a_{t}), \label{eq:word2vec-obj}
\end{equation}
where,

%
% CONDITIONAL PROBABILITY EQ
%
\begin{equation}
	P(j \given i) = \frac{\exp(\vect{u}_j \cdot \vect{v}_{i})}{Z_i}, \label{eq:cond_prob_w2v}
\end{equation}
where $\vect{v}$ and $\vect{u}$ are the ``in-vector'' and ``out-vector'', respectively,  $Z_i=\sum_{j' \in \mathcal{A}} \exp(\vect{u}_{j'} \cdot \vect{v}_{i})$ is a normalization constant, and $\mathcal{A}$ is the set of all locations. In general, calculating $Z_i$ is computationally expensive and there are two common approximation:
hierarchical softmax \autocite{morin2005hierarchical} and negative sampling \autocite{mikolov2013word2vec}.
Due to its simplicity and performance, negative sampling is the most widely used strategy, which we also adopt in our study.

\subsection{An implicit bias in the negative sampling}

Although negative sampling is the most common approximation, it is a biased estimator~\autocite{Chia2010,Dyer2014} that does not yield the optimal embeddings in terms of the objective function, ${\cal J}$.
This bias in negative sampling turns out to be the critical piece that connects word2vec and the gravity model.

Negative sampling trains {\it word2vec} using a binary classification task as follows.
For each target word $i$, we sample a context word $j$ from the given data and label it as positive, denoted by $Y_{j}=1$.
Then, we sample $k$ words $\ell$ from a noise distribution $p_0(\ell)$ and label them as negative, denoted by $Y_{\ell}=0$.
In \textit{word2vec}, the noise distribution is given by $p_0(\ell) \propto P^\gamma (\ell)$, where $P(j)$ is the fraction of $j$ in the data, and $\gamma$ is a hyper-parameter.
Then, for the sampled words, we fit a logistic regression model
\begin{align}
	\label{eq:logistic-regress}
	P^{\text{NS}}(Y_{j} = 1; \vect{v}_{i}, \vect{u}_j) = \frac{1}{1 + \exp(-\vect{u}_j \cdot \vect{v}_{i})},
\end{align}
by maximizing the log-likelihood:
\begin{align}
	\label{eq:log-likelihood-logistic-regress}
	{\cal J}^{\text{NS}} = \sum_{i \in {\cal A}} \sum_{j \in {\cal D}} \left[ Y_{j} \log P^{\text{NS}}(Y_{j} = 1; \vect{v}_{i}, \vect{u}_j) + (1-Y_{j}) \log P^{\text{NS}}(Y_{j} = 0; \vect{v}_{i}, \vect{u}_j)\right],
\end{align}
where ${\cal D}$ is the set of all sampled context words.

An underappreciated fact is that maximizing ${\cal J}^{\text{NS}}$ does not guarantee
to yield the optimum embedding in terms of the original skip-gram word2vec objective function ${\cal J}$ even if we increase the training samples and iterations~\autocite{Chia2010,Dyer2014}.
To make this bias explicit, let us consider the unbiased variant of negative sampling, \ie the noise contrastive estimation (NCE)~\autocite{Chia2010,Dyer2014}.
NCE is an unbiased estimator for a probability model $P_m$ of the form:
\begin{align}
	\label{eq:nce-model}
	P_m(x) = \frac{ f(x) }{\sum_{ x' \in {\cal X}} f(x') },
\end{align}
where $f$ is a non-negative likelihood function of data $x$, and ${\cal X}$ is the set of all data.
This general form includes the word2vec model (Eq.~\eqref{eq:cond_prob_w2v}), where $f(x) = \exp(x)$ and $x = \vect{u}_j \cdot \vect{v}_{i}$.
NCE fits the probability model using a binary classification task in the same way as in negative sampling but using a Bayesian formalism for logistic regression.
Specifically, before the training, we know that $1$ in $1+k$ words is sampled from the given data, which can be modeled as prior probabilities
\begin{align}
	\label{eq:prior}
	P(Y_{j} = 1) = \frac{1}{k + 1},\quad P(Y_{j} = 0) = \frac{k}{k + 1}.
\end{align}
Using the Bayes rule, the posterior probability for $Y_{j}$ given word $j$ is given by
\begin{align}
	\label{eq:posterior}
	P\left(Y_{j} \vert j\right) = \frac{
		P\left(j \vert Y_{j}\right)P(Y_{j})
	}{
		P\left(j \vert Y_{j} = 0\right)P(Y_{j} = 0)
		+ P\left(j \vert Y_{j} = 1\right)P(Y_{j} = 1)
	}.
\end{align}
Bearing in mind that word $j$ is sampled from the given data if $Y_{j}=1$ and from the noise distribution $p_0$ if $Y_j = 0$.
Assuming that the given data is generated from the probability model to fit, the class-conditional probability, $P\left(j \vert Y_{j}\right)$, is given by
\begin{align}
	\label{eq:class-cond}
	P(j \vert Y_{j} = 1) & = P_m(\vect{u}_j \cdot \vect{v}_{i}),
	\quad
	P(j \vert Y_{j} = 0)  = p_0 (j).
\end{align}
Putting Eqs.~\eqref{eq:prior}, \eqref{eq:posterior} and \eqref{eq:class-cond} together, the posterior probability for $Y_j$ is given by
\begin{align}
	\label{eq:nce}
	P\left(Y_{j} = 1 \vert j\right) & =
	\frac{
		P_m(\vect{u}_j \cdot \vect{v}_{i}) / (k + 1)
	}{
		P_m(\vect{u}_j \cdot \vect{v}_{i}) / (k + 1)  + kp_0(j) / (k + 1)
	}                                          \\
	                                & = \frac{
		P_m(\vect{u}_j \cdot \vect{v}_{i})
	}{
		P_m(\vect{u}_j \cdot \vect{v}_{i})  + kp_0(j)
	},
\end{align}
which can be rewritten in form of sigmoid function:
\begin{align}
	\label{eq:nce}
	P^{\text{NCE}}\left(Y_{j}=1 \vert j\right) =
	 & = \frac{
		1
	}{
		1 + kp_0(j) / P_m(\vect{u}_j \cdot \vect{v}_{i})
	}           \\
	 & = \frac{
		1
	}{
		1 + \exp\left[ \ln kp_0(j) - \ln P_m(\vect{u}_j \cdot \vect{v}_{i}) \right]
	}           \\
	 & = \frac{
		1
	}{
		1 + \exp\left[ - \ln f(\vect{u}_j \cdot \vect{v}_{i})  + \ln p_0(j) + c \right]
	},
\end{align}
where $c = \ln k + \ln\sum_{ x' \in {\cal X}} f(x') $ is a constant.
NCE maximizes the log-likelihood
\begin{align}
	\label{eq:log-likelihood-logistic-regress}
	{\cal J}^{\text{NCE}} = \sum_{i \in {\cal A}}\sum_{j \in {\cal D}} \left[ Y_{j} \log P^{\text{NCE}}(Y_{j} = 1\vert j) + (1-Y_{j}) \log P^{\text{NCE}}(Y_{j} = 0 \vert j)\right].
\end{align}
by calculating the gradients for embedding vectors $\vect{u}_j$, $\vect{v}_{i}$ and iteratively updating them.
An important consequence of this framework is that NCE is an unbiased estimator that has convergence to the optimal embedding in terms of the original word2vec's objective function, ${\cal J}$~\autocite{Chia2010,Dyer2014} if we increase the number of words to sample and the training iterations.
